Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-05-22 manuscript.qmd
pushing a first version of intro+methods+results

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window into the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it makes difficult to draw precise conclusions about the onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our approach leverages Bayesian generalised additive multilevel models, providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# see available font via systemfonts::system_fonts()
# monofont: Monaco
# monofont: Courier
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
suppress-title-introduction: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
          \usepackage{setspace}
          \onehalfspacing
---

@@

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography and electroencephalography (M/EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether, when, and where neural responses differ across experimental conditions or participant groups.

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form or correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or the false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab`, @delorme2004; `MNE-Python`,  @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it prevents the ability to draw precise conclusions about the spatiotemporal localisation of such effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007: "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for localising the onset of M/EEG effects; a property that was later demonstrated in simulations studies [e.g., @rousselet_using_2025; @sassenhagen2019].

@@

The issues with multiple comparisons represent a common and well-recognised danger in neuroimaging and M/EEG research where the collected data allows for a multitude of potential hypothesis tests and is characterised by complex structures of spatiotemporal dependencies. The probability of obtaining at least one false positive in an ensemble (family) of $m$ tests (i.e., the FWER) is computed as $1-\left(1-\alpha\right)^{m}$ (for $m=10$ tests and $\alpha=0.05$, it is approximately equal to $0.4$). Different methods exist to control the FWER (i.e., bring it back to $\alpha$). Most methods apply a simple correction to series of "raw" $p$-values issued from univariate statistical tests (e.g., t-tests). For instance, the Bonferroni correction [@dunn1961] consists in setting the significance threshold to $\alpha/m$, or equivalently, multiplying the "raw" $p$-values by $m$ and using the standard $\alpha$ significance threshold. This method is generally overconservative (i.e., under-powered) as it assumes statistical independence of the tests, an assumption that is clearly violated in the context of M/EEG timeseries characterised by massive spatiotemporal dependencies. Some alternative methods aims at controlling the FDR, defined as the proportion of false positive *among positive tests* [e.g., @benjamini1995; @benjamini2001]. However, a major limitation of both types of corrections is that they do not take into account the spatial and temporal information contained in M/EEG data.

A popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference [@bullmore1999; @maris2007]. A typical cluster-based inference consists of two successive steps [for more details on cluster-based inference, see for instance @frossard2022; @maris2011; @maris2007; @sassenhagen2019]. First, clusters are defined as sets of contiguous timesteps, sensors, voxels, etc, whose activity, summarised by some test statistic (e.g., a $t$-value), exceeds a predefined threshold (e.g., the 95th percentile of the parametric null distribution). Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, for instance by summing the statistics within a cluster, an approach referred to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by computing a $p$-value for each identified cluster by comparing its mass with the null distribution of cluster masses (obtained via permutation). As alluded previously, a significant cluster is a cluster which contains *at least one* significant time-point. As such, it would be incorrect to conclude, for instance, that the timestep of a significant cluster is the first moment at which some conditions differ [@frossard2022; @sassenhagen2019]. In other words, because the inference is performed at the second step (i.e., once clusters have been formed), it prevents any conclusion to be made about individual datapoints (e.g., timesteps, sensors, etc).

@@

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Note that in practise, this integral is approximated by a sum over small $h$ increments. Then, a p-value for each timestep $t$ is computed by comparing its TFCE with the null distribution of TFCE values (obtained via permutation). For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate FWER control [e.g., @pernet2015]. However, further simulation work showed that cluster-based methods (including TFCE) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

To sum up, the main limitation of cluster-based inference is that it allows for inference at the cluster level only, not allowing inference at the level of timesteps, sensors, etc. As a consequence, it does not allow inferring the precise spatial and temporal localisation of effects. In the following, we briefly review previous M/EEG modelling work. Then, we provide a short introduction to generalised additive models (GAMs) and Bayesian generalised additive multilevel models (BGAMMs) to illustrate how these models can be used to precisely localise the onset and offset of M/EEG effects.

@@

Recent example of GLM for EEG [@wüllhorst2025; @fischer2013]... See also [@rousselet2008; @hauk2006]... Example of two-stage regression analysis [i.e., individual-level then group-level, @dunagan2024]...

See also the rERP framework [@smith2014a; @smith2014b] and @tremblay2014...

From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]... GAMMs for EEG data [@abugaber2023; @meulman2015; @meulman2023]... See also the UNFOLD toolbox [@ehinger_unfold_2019]...

Disentangling overlapping processes [@skukies_brain_2024; @skukies_modelling_2021]... Weighting single trials [@pernet2022]... The LIMO toolbox [@pernet2011] using linear and 2-stage regression (not a proper multilevel model)...

Recently, @teichmann2022 provided a detailed tutorial on using Bayes factors (BFs) to analyse the 1D or 2D output from MVPA, that is, for testing, at every timestep, whether decoding performance is above chance level. However, this approach provides timeseries of BFs that ignores temporal dependencies...

@@

See for instance these tutorials [@sóskuthy2017; @winter2016] or application to phonetic data [@wieling2018; @sóskuthy2021] or this introduction [@baayen2020] or these reference books [@hastie2017; @wood2017]... application to pupillometry [@vanrij2019]... GAMLSS for neuroimaging data [@dinga2021]... Modelling auto-correlation in GAMMs + EEG example [@baayen2018]...

In generalised additive models (GAMs), the functional relationship between predictors and response variable is decomposed into a sum of low-dimensional non-parametric functions. A typical GAM has the following form:

@@

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $\mathbf{A}_{i}$ is the $i$th row of a known parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms (to be estimated), $f_{j}$ is a smooth function of covariate $x_{j}$ (to be estimated as well). The smooth functions $f_{j}$ are represented in the model via penalised splines basis expansions of the covariates, that are a weighted sum of $K$ simpler, basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this stage: *splines* are functions composed of simpler functions. These simpler functions are basis functions (e.g., cubic polynomial, thin-plate) and the set of basis functions is a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function. In other words, more complex (wiggly) basis functions are penalised.

@@

The Bayesian approach to statistical modelling is characterised by its reliance on probability theory to [@gelman2020]... In this framework, all unknown entities are assigned probability distributions reflecting the uncertainty... These probability distributions are commonly referred to as "priors" and represent some state of knowledge about unknown quantities before seeing any data. There are debates among Bayesian practitioners as to whether prior distributions should encore subjective (personal) beliefs or... but these debates are outside the scope of the present paper and we therefore the interested reader to dedicated work (e.g., XX; YY)... In practice, weakly informative priors are often used as default priors in situations in which subjective priors are difficult to define/elicit... Bayesian models are then fitted on empirical (actual or simulated) data to update prior states of knowledge to posterior states of knowledge using Bayes theorem, or in practise, sampling-based approximations of the posterior distribution...

```{r fig-intro-gam, echo = FALSE, fig.width = 9, fig.cap = "Different types of GAM(M)s. **A**: GAMs predictions are computed as the weigthed sum (in black) of basis functions (here thin-plate basis functions, in grey). **B**: Constant-effect GAM, with 5 participants in colours and the group-level prediction in black. **C**: Varying-intercept + varying-slope GAMM (with common smoother). **D**: Varying-intercept + varying-slope + varying-smoother GAMM. In this model, each participant gets its own intercept, slope, and degree of 'wiggliness' (smoother)."}
########################
# simulating som edata #
########################

@@

    # plotting everything
    ggplot() +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_line(
            # data = basis_long, aes(x = x, y = value, color = basis),
            data = basis_long, aes(x = x, y = value, group = basis),
            colour = "grey50",
            alpha = 0.5,
            linewidth = 0.5
            ) +
        geom_line(
            data = df,
            aes(x = x, y = fit),
            color = "black",
            linewidth = 1.0
            ) +
        # geom_text(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 0.1, size = 2
        #     ) +
        # ggrepel::geom_text_repel(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 1,
        #     size = 2, direction = "y"
        #     ) +
        labs(x = "Predictor", y = "Response") +
        ylim(-2, +2) +
        # scale_colour_manual(values = met.brewer(name = "Johnson", n = 10) ) +
        theme(legend.position = "none")

@@

    # plotting the predictions
    df %>%
        ggplot(aes(x = x, y = y, colour = subject) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            size = 0.5,
            show.legend = FALSE
            ) +
        geom_line(
            aes(y = fit_subject, group = subject),
            alpha = 0.5,
            linewidth = 0.5,
            show.legend = FALSE
            ) +
        geom_ribbon(
            aes(x = x, y = y, ymin = fit_global_lower, ymax = fit_global_upper),
            fill = "black",
            alpha = 0.2,
            inherit.aes = FALSE
            ) +
        geom_line(
            aes(y = fit_global, group = 1),
            color = "black",
            linewidth = 1
            ) +
        ylim(-5, 5) +
        scale_colour_manual(values = met.brewer(name = "Johnson", n = 5) ) +
        labs(x = "Predictor", y = "Response")

@@

# plotting, note that the basis plot uses the fixed-effect model for simplicity
p1 <- plot_basis_functions(df, gam_fixed)
p2 <- plot_model(df, gam_fixed)
p3 <- plot_model(df, gam_int_slope)
p4 <- plot_model(df, gam_rand_smooth)

@@

See @fig-intro-gam... Introduction to multilevel GAMs [@pedersen_hierarchical_2019]... Now describe the Bayesian GAMM [@miller2025]... Proper inclusion of varying/random effects in the model specification protects against overly wiggly curves [@baayen2020]... Generalising to scale and shape or "distributional GAMs" [@rigby2005; @umlauf2018] and applied to neuroimaging data [@dinga2021]...

>>>>

---
title: "Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models"
shorttitle: "Bayesian modelling of M/EEG data"
author:
  - name: Ladislas Nalborczyk
    corresponding: true
    orcid: 0000-0002-7419-9855
    email: ladislas.nalborczyk@cnrs.fr
    url: https://lnalborczyk.github.io
    affiliations:
      - id: id1
        name: "Aix Marseille Univ, CNRS, LPL"
        address: "5 avenue Pasteur"
        city: "13100 Aix-en-Provence"
        country: France
  - name: Paul Bürkner
    orcid: 0000-0001-5765-8995
    url: https://paulbuerkner.com
    affiliations:
      - name: "TU Dortmund University, Department of Statistics"
author-note:
  disclosures:
    conflict-of-interest: "The authors have no conflicts of interest to disclose."
abstract: "Time-resolved electrophysiological measurements such as those obtained through magneto- or electro-encephalography (M/EEG) offer a unique window onto the neural activity underlying cognitive processes. Researchers are often interested in determining whether and when these signals differ across experimental conditions or participant groups. The conventional approach involves mass-univariate statistical testing across time and space, followed by corrections for multiple comparisons such as cluster-based inference. While effective for controlling error rates at the cluster-level, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual time points to clusters, it prevents drawing conclusions about the precise onset or offset of observed effects. Here, we introduce a *model-based* approach for analysing M/EEG timeseries such as event-related potentials (ERPs) or decoding performance over time. Our approach leverages Bayesian generalised additive multilevel models, providing posterior probabilities that an effect is above zero (or above chance) at each time point, while naturally accounting for temporal dependencies and between-subject variability. Using both simulated and actual M/EEG datasets, we demonstrate that this approach substantially outperforms conventional methods in estimating the onset and offset of neural effects, yielding more precise and reliable results. We provide an R package implementing the method and describe how it can be integrated into M/EEG analysis pipelines using MNE-Python."
keywords: [EEG, MEG, cluster-based inference, simulation, multiple comparisons, generalised additive models, mixed-effects models, multilevel models, Bayesian statistics, brms]
floatsintext: true
numbered-lines: true
bibliography:
  - meeg_modelling.bib
  - grateful-refs.bib
suppress-title-page: false
link-citations: true
# see https://www.overleaf.com/learn/latex/Font_typefaces
# see available font via systemfonts::system_fonts()
# monofont: Monaco
# monofont: Courier
mainfont: CMU Serif
fontsize: 10pt
draftfirst: false
draftall: false
a4paper: true
donotrepeattitle: true
suppress-title-introduction: true
lang: en
toc: true
language:
  citation-last-author-separator: "&"
  email: "email"
  title-block-author-note: "Author note"
  title-block-correspondence-note: "Correspondence concerning this article should be addressed to"
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows:"
format:
  apaquarto-pdf:
    latex-engine: xelatex
    documentmode: doc # man, jou, doc, stu
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{mathtools}
          \usepackage{booktabs, caption, longtable, colortbl, array}
          \usepackage{setspace}
          \onehalfspacing
---

@@

# disabling hyperlinks
# options(cli.hyperlink = FALSE)

@@

Understanding the temporal dynamics of cognitive processes requires methods that can capture fast-changing neural activity with high temporal resolution. Magnetoencephalography and electroencephalography (M/EEG) are two such methods, widely used in cognitive neuroscience for their ability to track brain activity at the millisecond scale. These techniques provide rich time series data that reflect how neural responses unfold in response to stimuli or tasks. A central goal in many M/EEG studies is to determine whether, when, and where neural responses differ across experimental conditions or groups.

The conventional approach involves mass-univariate statistical testing through time and/or space followed by some form of correction for multiple comparisons with the goal of maintaining the familywise error rate (FWER) or the false discovery rate (FDR) at the nominal level (e.g., 5%). Cluster-based inference is the most common way of achieving this sort of error control in the M/EEG literature, being the recommended approach in several software programs [e.g., `EEGlab`, @delorme2004; `MNE-Python`, @gramfort2013]. While effective for controlling error rates, cluster-based inference comes with a significant limitation: by shifting the focus of inference from individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it prevents the ability to draw precise conclusions about the spatiotemporal localisation of such effects [@maris2007; @sassenhagen2019]. As pointed out by @maris2007: "there is a conflict between this interest in localized effects and our choice for a global null hypothesis: by controlling the FA [false alarm] rate under this global null hypothesis, one cannot quantify the uncertainty in the spatiotemporal localization of the effect". Even worse, @rosenblatt2018 note that cluster-based inference suffers from low spatial resolution: "Since discovering a cluster means that 'there exists at least one voxel with an evoked response in the cluster', and not that 'all the voxels in the cluster have an evoked response', it follows that the larger the detected cluster, the less information we have on the location of the activation." As a consequence, cluster-based inference is expected to perform poorly for localising the onset of M/EEG effects; a property that was later demonstrated in simulation studies [e.g., @rousselet_using_2025; @sassenhagen2019].

@@

The issues with multiple comparisons represent a common and well-recognised danger in neuroimaging and M/EEG research, where the collected data allows for a multitude of potential hypothesis tests and is characterised by complex structures of spatiotemporal dependencies. The probability of obtaining at least one false positive in an ensemble (family) of $m$ tests (i.e., the FWER) is computed as $1-\left(1-\alpha\right)^{m}$ (for $m=10$ tests and $\alpha=0.05$, it is approximately equal to $0.4$). Different methods exist to control the FWER (i.e., bring it back to $\alpha$). Most methods apply a simple correction to series of $p$-values issued from univariate statistical tests (e.g., t-tests). For instance, the Bonferroni correction [@dunn1961] consists in setting the significance threshold to $\alpha/m$, or equivalently, multiplying the $p$-values by $m$ and using the standard $\alpha$ significance threshold. This method is generally overconservative (i.e., under-powered) as it assumes statistical independence of the tests, an assumption that is clearly violated in the context of M/EEG timeseries characterised by massive spatiotemporal dependencies. Some alternative methods aims at controlling the FDR, defined as the proportion of false positive *among positive tests* [e.g., @benjamini1995; @benjamini2001]. However, a major limitation of both types of corrections is that they do not take into account the spatial and temporal information contained in M/EEG data.

A popular technique to account for spatiotemporal dependencies while controlling the FWER is cluster-based inference [@bullmore1999; @maris2007]. A typical cluster-based inference consists of two successive steps [for more details on cluster-based inference, see for instance @frossard2022; @maris2011; @maris2007; @sassenhagen2019]. First, clusters are defined as sets of contiguous timesteps, sensors, voxels, etc, whose activity, summarised by some test statistic (e.g., a $t$-value), exceeds a predefined threshold (e.g., the 95th percentile of the parametric null distribution). Clusters are then characterised by their height (i.e., maximal value), extent (number of constituent elements), or some combination of both, for instance by summing the statistics within a cluster, an approach referred to as "cluster mass" [@maris2007; @pernet2015]. Then, the null hypothesis is tested by computing a $p$-value for each identified cluster by comparing its mass with the null distribution of cluster masses (obtained via permutation). As alluded previously, a significant cluster is a cluster which contains *at least one* significant time-point. As such, it would be incorrect to conclude, for instance, that the timestep of a significant cluster is the first moment at which some conditions differ [@frossard2022; @sassenhagen2019]. In other words, because the inference is performed at the second step (i.e., once clusters have been formed), no conclusion can be made about individual datapoints (e.g., timesteps, sensors, etc).

@@

where $h_{0}$ is typically $0$ and parameters $E$ and $H$ are set a priori (typically to 0.5 and 2, respectively) and control the influence of the extend and height on the TFCE. Note that in practice, this integral is approximated by a sum over small $h$ increments. Then, a *p*-value for each timestep $t$ is computed by comparing its TFCE with the null distribution of TFCE values (obtained via permutation). For each permuted signal, we keep the maximal value over the whole signal for the null distribution of the TFCE. The TFCE combined with permutation (assuming a large enough number of permutations) has been shown to provide accurate FWER control [e.g., @pernet2015]. However, further simulation work showed that cluster-based methods (including TFCE) perform poorly in localising the onset of M/EEG effects [e.g., @rousselet_using_2025; @sassenhagen2019].

To sum up, the main limitation of cluster-based inference is that it allows for inference at the cluster level only, not allowing inference at the level of timesteps, sensors, etc. As a consequence, it does not allow inferring the precise spatial and temporal localisation of effects. In the following, we briefly review previous modelling work of M/EEG data. Then, we provide a short introduction to generalised additive models (GAMs) to illustrate how these models can be used to precisely estimate the onset and offset of M/EEG effects.

@@

Weighting single trials [@pernet2022]... From @dimigen2021: Recently, spline regression has been applied to ERPs [e.g., @hendrix2017; @kryuchkova2012]...

M/EEG recordings at the scalp measure activity stemming from various parts of the brain and contain many artefacts not related to the cognitive and neural processes of interest. Therefore, analysing M/EEG data requires methods for separating the parts of the signals that are related to cognitive and neural processes and the parts of the signals that contain task-unrelated "noise". The most widely-used method for achieving this separation consists in estimating the event-related potential (ERP), defined as the brain response elicited by some specific time-locked sensory, cognitive, or motor event. Most often, the ERP is computed as the average response over trials computed from epochs time-locked on the event of interest (e.g., stimulus onset). However, this simple averaging approach comes with several limitations, such as being limited to simple categorical designs.

Scalp-recorded M/EEG signals capture neural activity originating from various brain regions and are often contaminated by artifacts unrelated to the cognitive processes under investigation. Consequently, analysing M/EEG data necessitates methods that can disentangle task-relevant neural signals from extraneous "noise." A widely adopted technique for this purpose is the estimation of event-related potentials (ERPs), which are stereotyped electrophysiological responses time-locked to specific sensory, cognitive, or motor events. Typically, ERPs are derived by averaging EEG or MEG epochs across multiple trials aligned to the event of interest (e.g., stimulus onset), thereby enhancing the signal-to-noise ratio by attenuating non-time-locked activity. However, this averaging approach has notable limitations: it assumes consistent latency and amplitude across trials and is primarily suited for simple categorical designs. Such assumptions may not hold in more complex experimental paradigms, potentially leading to suboptimal ERP estimations.

To overcome the limitations of averaging, a number of model-based options for estimating ERPs have been proposed. These approaches are justified by the observation that the averaging approach to ERPs is equivalent to an intercept-only linear regression in a simple categorical design with no overlapping factors [@smith2014a]. The regression approach to ERPs is much more flexible than simpl averaging, as it allows modelling linear and non-linear effects of continuous predictors such as word frequency or age [e.g., @smith2014a; @smith2014b; @tremblay2014] and allow modelling and disentangling overlapping processes [e.g., @skukies_brain_2024; @skukies_modelling_2021] as implemented in the UNFOLD toolbox [@ehinger_unfold_2019]. The regression approach, as implemented for instance in the LIMO toolbox [@pernet2011], is a multi-stage procedure in which a participant-level regression model is fitted independently to each datapoint (e.g., channel, timestep, etc.) to estimate the ERP, followed by group-level statistical analyses of the regression coefficients, itself followed by multiplicity corrections or cluster-based inference [for recent applied examples, see for instance @dunagan2025; @wüllhorst2025]. Whereas this approach allows including linear and non-linear discrete and continuous predictors when estimating the ERP, it has several limitations. First, the participant-level modelling ignores spatiotemporal dependencies as it treats all datapoints independently. Second, this approach does not take into account the participant (or stimulus) variability when testing group-level hypotheses (as could be done for instance via multilevel modelling). Finally, because the output of this procedure is summarised by cluster-based inference, its conclusions suffer from the same limitations discussed in the previous section.

To overcome the limitations of simple averaging, several model-based approaches for estimating ERPs have been proposed. These methods are motivated by the observation that traditional ERP averaging is mathematically equivalent to fitting an intercept-only linear regression model in a simple categorical design without overlapping events [@smith2014a]. In contrast to simple averaging, regression-based approaches to ERP estimation offer substantially greater flexibility. Notably, they allow for the modelling of both linear and nonlinear effects of continuous predictors--such as word frequency or age [e.g., @smith2014a; @smith2014b; @tremblay2014]--and enable the disentangling of overlapping cognitive processes [e.g., @ehinger_unfold_2019; @skukies_brain_2024; @skukies_modelling_2021]. One widely used implementation of this approach is provided by the LIMO EEG toolbox [@pernet2011], which follows a multi-stage analysis pipeline. First, a separate regression model is fit for each datapoint (e.g., each time point and electrode) at the individual participant level to estimate ERP responses. This is followed by group-level statistical analyses of the resulting regression coefficients, often accompanied by multiple comparison corrections or cluster-based inference [for recent applied examples, see @dunagan2025; @wüllhorst2025].

Although this framework allows for the inclusion of a wide range of predictors--both continuous and categorical, linear and nonlinear--it still has important limitations. First, fitting separate models for each datapoint ignores the spatiotemporal dependencies inherent in M/EEG data, potentially reducing statistical power and interpretability. Second, the subsequent group-level analyses typically do not account for hierarchical dependencies, between-participant or between-stimulus variability, which could otherwise be addressed through multilevel modelling. Finally, because the output of this procedure is summarised by cluster-based inference, its conclusions remain subject to the limitations discussed in the previous section.

Building upon the regression framework, recent advances allow modelling and disentangling overlapping processes [e.g., @skukies_brain_2024; @skukies_modelling_2021] as implemented in the UNFOLD toolbox [@ehinger_unfold_2019]...

In addition to modelling the non-linear effects of continuous predictors (e.g., word frequency) on the ERP amplitude, GAMs have been used for modelling the effect of time, that is, the shape of the ERP itself [@abugaber2023; @baayen2018; @meulman2015; @meulman2023]... In the next section, we turn to a brief introduction to generalised additive models and discuss their use for modelling M/EEG timeseries...

Beyond modelling nonlinear effects of continuous predictors on ERP amplitudes, generalised additive models (GAMs) have been employed to capture the temporal dynamics of ERPs themselves, effectively modelling the shape of the waveform over time [@abugaber2023; @baayen2018; @meulman2015; @meulman2023]. This approach allows for the estimation of smooth, data-driven functions that characterise how neural responses evolve over time, offering a flexible alternative to traditional linear models. In the following section, we provide a brief introduction to GAMs, highlighting their applicability to M/EEG time series analysis and the advantages they offer over conventional methods.

@@

In generalised additive models (GAMs), the functional relationship between the predictors and the response variable is decomposed into a sum of low-dimensional non-parametric functions. A typical GAM has the following form:

@@

where $y_{i} \sim \mathrm{EF}\left(\mu_{i}, \phi\right)$ denotes that the observations $y_{i}$ are distributed as some member of the exponential family of distributions (e.g., Gaussian, Gamma, Beta, Poisson) with mean $\mu_{i}$ and scale parameter $\phi$; $g(\cdot)$ is the link function, $\mathbf{A}_{i}$ is the $i$th row of a known parametric model matrix, $\gamma$ is a vector of parameters for the parametric terms (to be estimated), $f_{j}$ is a smooth function of covariate $x_{j}$ (to be estimated as well). The smooth functions $f_{j}$ are represented in the model as a weighted sum of $K$ simpler, basis functions:

@@

where $\beta_{jk}$ is the weight (coefficient) associated with the $k$th basis function $b_{jk}()$ evaluated at the covariate value $x_{ij}$ for the $j$th smooth function $f_{j}$. To clarify the terminology at this point: *splines* are functions composed of simpler functions. These simpler functions are called *basis functions* (e.g., cubic polynomial, thin-plate) and the set of basis functions is called a *basis*. Each basis function gets its coefficient and the resultant spline is the sum of these weighted basis functions (@fig-intro-gam\ignorespaces A). Splines coefficients are penalised (usually through the squared of the smooth functions' second derivative) in a way that can be interpreted, in Bayesian terms, as a prior on the "wiggliness" of the function [@miller2025; @wood2017]. In other words, more complex (wiggly) basis functions are automatically penalised.

A detailed discussion of the technical aspects of smooth functions is beyond the scope of this article (see Wood, 2017a); or (Baayen, Vasishth, Kliegl, & Bates, 2017; van Rij, Vaci, Wurm, & Feldman, in press; Wieling, 2018, for an introduction), but in the context of this work, it can be thought of as a continuous, potentially wiggly but not abruptly changing line that is expressed over time.

Generalised additive models can be understood in a Bayesian framework [@miller2025; @wood2017]...

@@

There are debates among Bayesian practitioners as to whether prior distributions should encore subjective (personal) beliefs or... but these debates are outside the scope of the present paper and we therefore the interested reader to dedicated work (e.g., XX; YY)... In practice, weakly informative priors are often used as default priors in situations in which subjective priors are difficult to define/elicit...

The Bayesian approach to statistical modelling is characterised by its reliance on probability theory to quantify uncertainty about parameter estimates [e.g., @gelman2020]. In this framework, all unknown entities are assigned probability distributions reflecting the uncertainty... These probability distributions are commonly referred to as "priors" and represent some state of knowledge about unknown quantities before seeing any data. Bayesian models are then fitted on empirical (actual or simulated) data to update prior states of knowledge to posterior states of knowledge using Bayes theorem, or in practise, sampling-based approximations of the posterior distribution...

```{r fig-intro-gam, echo = FALSE, fig.width = 9, fig.cap = "Different types of GAMs. **A**: GAMs predictions are computed as the weigthed sum (in black) of basis functions (here thin-plate basis functions, in grey). **B**: Constant-effect GAM, with 5 participants in colours and the group-level prediction in black. **C**: Varying-intercept + varying-slope GAMM (with constant smoother). **D**: Varying-intercept + varying-slope + varying-smoother GAMM. In this model, each participant gets its own intercept, slope, and degree of 'wiggliness' (smoother)."}
########################
# simulating some data #
########################

@@

    # plotting everything
    ggplot() +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_line(
            # data = basis_long, aes(x = x, y = value, color = basis),
            data = basis_long, aes(x = x, y = value, group = basis),
            colour = "grey50",
            alpha = 0.5,
            linewidth = 0.5
            ) +
        geom_line(
            data = df,
            aes(x = x, y = fit),
            color = "black",
            linewidth = 1.0
            ) +
        # geom_text(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 0.1, size = 2
        #     ) +
        # ggrepel::geom_text_repel(
        #     data = label_df,
        #     aes(x = x, y = value, label = label, color = basis),
        #     hjust = 0, nudge_x = 1,
        #     size = 2, direction = "y"
        #     ) +
        labs(x = "Predictor", y = "Response") +
        ylim(-2, +2) +
        theme(legend.position = "none")

@@

    # plotting the predictions
    df %>%
        ggplot(aes(x = x, y = y, colour = subject) ) +
        geom_hline(yintercept = 0, linetype = 2) +
        geom_point(
            size = 0.5,
            show.legend = FALSE
            ) +
        geom_line(
            aes(y = fit_subject, group = subject),
            alpha = 0.5,
            linewidth = 0.5,
            show.legend = FALSE
            ) +
        geom_ribbon(
            aes(x = x, y = y, ymin = fit_global_lower, ymax = fit_global_upper),
            fill = "black",
            alpha = 0.2,
            inherit.aes = FALSE
            ) +
        geom_line(
            aes(y = fit_global, group = 1),
            color = "black",
            linewidth = 1
            ) +
        ylim(-5, 5) +
        scale_colour_manual(values = met.brewer(name = "Johnson", n = n_subjects) ) +
        labs(x = "Predictor", y = "Response")

@@

# plotting, note that the basis plot uses the fixed-effect model for simplicity
# df2 <- df %>% filter(subject %in% c(unique(df$subject)[1:4]) )
p1 <- plot_basis_functions(df, gam_fixed)
p2 <- plot_model(df, gam_fixed)
p3 <- plot_model(df, gam_int_slope)
p4 <- plot_model(df, gam_rand_smooth)

@@

Or (Baayen, Vasishth, Kliegl, & Bates, 2017; van Rij, Vaci, Wurm, & Feldman, in press; Wieling, 2018, for an introduction), but in the context of this work, it can be thought of as a continuous, potentially wiggly but not abruptly changing line that is expressed over time.
